---
title: "Machine Learning on Accelerometer Data"
author: "Reynaldo"
date: "5/9/2017"
output: 
  html_document: 
    self_contained: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning=FALSE, message=FALSE, cache = TRUE,
                      fig.align = 'center')
```

<br>

**UPDATE:** New and adjusted specifications on the original dataset attain **accuracies of over 99.9%**, with precision, recall, and F1 of over 99.9% as well on all classes. See the analysis in this  [Jupyter Notebook](http://nbviewer.jupyter.org/github/reyvaz/Qualifying-Human-Activity/blob/master/qualifying_workouts.ipynb), or check out the repo [here](https://github.com/reyvaz/Qualifying-Human-Activity).


## Summary  
This analysis uses different machine learning algorithms on accelerometer data to predict the way individuals perform weight-lifting exercises. The dataset comes from Veloso et al., (2013) and it contains data from accelerometers on the belt, forearm, arm, and dumbbell from 6 individuals.

The individuals were asked to perform one set of 10 repetitions of the unilateral dumbbell biceps curl in five different ways: (A) correctly; (B) throwing the elbows to the front; (C) lifting the dumbbell only halfway; (D) lowering the dumbbell only halfway; and (E) throwing the hips to the front.  

The following analysis tests different machine learning algorithms, including CART, Random Forest, and Boosted (GBM) to predict the way dumbbell biceps curls were performed.  The best performing algorithm is a Random Forest specification with $99.3\%$ accuracy, followed by a Boosted GBM with $94.3\%$ accuracy, in the test dataset. 

1. Necessary packages
```{r packs}
library(caret); library(dplyr); library(knitr); library(pander)
library(rpart); library(rpart.plot); library(gbm)
library(randomForest); library(ggRandomForests); library(corrplot)
```
## Data Processing  
The dataset is available [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv).
(Note the original name of the file is pml-training.csv. This is the entire data set that will be used for the following analysis. This pml-training.csv dataset will be split into an actual training-set and a testing-set)

2. Data
```{r data, results = "hide"}
data <- read.csv("pml-training.csv")
```
Preliminary analysis found: (a) a large number of variables with near zero variability; (b) the first columns contain recording and identification data irrelevant to the prediction; and (c) a significant number of variables with less than $10\%$ of valid observations. The variables with these characteristics are thus discarded.  

3. Discard variables    
        a. discard variables with near zero variance      
        b. discard variables irrelevant to prediction, columns 1-6      
        c. discard variables with 80% NAs or more
```{r discard, results = "hide"}
nrzv  <- nearZeroVar(data)
train <- data[,-nrzv]
train <- data[, -c(1:6)]
xNAs  <- which(colMeans(is.na(data)) > .8)
train <- data[, -xNAs]
```

**Dataset Split**  
Data is then split into a training dataset with 60% of observations for model training, and a testing dataset with the remaining 40% of observations.   

4. Split data for training and testing
```{r split, results = "hide"}
set.seed(400)
inTrain  = createDataPartition(data$classe, p = 3/5)[[1]]
training = data[ inTrain,]
testing  = data[-inTrain,]
```

Although correlation analysis shows a high correlation between a number of predictors (as seen in the visualization in the appendix), no further manual pre-processing will be performed for the remaining of the analysis. As a note, PCA decomposition (discarded in this report) was able to capture $95\%$ of the variability by reducing the number of components by over $50\%$, but accuracy was significantly compromised, while expediency gains were only minor. Because of the nature of decision trees (i.e. they can branch at equivalent splitting points) scaling or translational normalization is not necessary. Furthermore, they are also robust to correlated variates.  

## Machine Learning Specifications  

The following will test and compare the performance of 3 different machine learning algorithms: CART, Random Forest, and GBM. 

5. First Model: Classification and Regression Tree (CART)
```{r rcart}
tm.1      <- system.time(
rpart1    <- rpart(classe ~ ., method="class", data=training))
testrpart <- predict(rpart1, newdata = testing[,-length(testing)], type = "class")
cm1       <- confusionMatrix(testrpart, testing$classe)
```

6. Second Model: Random Forest (RF)
```{r rf, results = "hide"}
tm.2    <- system.time(
rfm     <- randomForest(training[,-length(training)], training[,length(training)], ntree = 500)    )
testrfm <- predict(rfm, newdata = testing[,-length(testing)])
cm2     <- confusionMatrix(testrfm, testing$classe)
```

7. Third Model: Boosting (GBM)
```{r gbm, results = "hide"}
tm.3      <- system.time(
gbm1      <- gbm.fit(x = training[,-length(training)], y = training[,length(training)],
                        distribution = "multinomial", verbose = FALSE, 
                        interaction.depth=5, shrinkage=0.005, n.trees = 1000))
best.iter <- gbm.perf(gbm1,method="OOB", plot.it = FALSE)
probs     <- predict(gbm1, testing[,-length(testing)], n.trees = best.iter, type = "response")
indexes   <- apply(probs, 1, which.max)
testgbm   <- colnames(probs)[indexes]
cm3       <- confusionMatrix(testgbm, testing$classe)
```

## Algorithm Performance Comparison

The following are some extractions from the Confusion Matrix output for each specification. These calculations were done on the test dataset (except time elapsed).

8. Performance Comparison
```{r Performance}
Accuracy     <- as.numeric(c(cm1$overall[1], cm2$overall[1], cm3$overall[1]))
Kappa        <- as.numeric(c(cm1$overall[2], cm2$overall[2], cm3$overall[2]))
OOBError     <- 1 - Accuracy
Time.Elapsed <- as.numeric(c(tm.1[3], tm.2[3], tm.3[3]))
Results      <- rbind(Accuracy, Kappa, OOBError, Time.Elapsed)
colnames(Results) <- c("CART", "Random Forest", "Boosted (GBM)")
```

#### The calculated Accuracy rates, Kappas, and Out of Sample Error rates estimates for each specification are:  

```{r Performance2, echo = FALSE}
kable(list(Results), caption = "**Model Performance**", digits = 4)
```

Performance metrics indicate that the Random Forest is the best performing algorithm for this purpose with an accuracy rate of $`r cm2$overall[1]`$. It is followed by the GBM algorithm with an accuracy rate of $`r cm3$overall[1]`$. And last, is the CART algorithm which performed poorly compared to the other 2 with an accuracy rate of $`r cm1$overall[1]`$. Although the CART algorithm can be further tuned to improve its performance, it is unlike to overperform the Random Forest. In terms of time efficiency, the Random Forest specification is about 7 times faster than GBM.   

The following reports the Confusion Matrices for the 2 best performing algorithms in terms of accuracy. Class Specific statistics are for the Random Forest model only.   

```{r CMs, echo = FALSE}
pander(cm3$table, caption = "**Confusion Matrix (GMB Model)**")
pander(cm2$table, caption = "**Confusion Matrix (Random Forest Model)**")
```


```{r Stats, echo = FALSE}
kable(list(t(cm2$byClass[,-c(5:7)])), caption = "**Statistics By Class (Random Forest Model)**", digits = 4)
```
  
## Conclusion:   
This analysis used accelerometer data to predict how well individuals perform dumbbell-lifting exercises. Three machine-learning algorithms were tested: CART, Random Forest, and GBM. The best performing algorithm was the Random Forest with $99.3\%$ accuracy, followed by the GBM with $94.3\%$ accuracy, and worst performing was the CART algorithm, which performed poorly compared to the other two with a $72.4\%$ accuracy.

## Appendix
Fig 1. Correlation Matrix Visualization
```{r plotcorr, fig.width=7.5, fig.height=7.5}
par(xpd=TRUE)
corrplot.mixed(cor(training[,-length(training)]), lower="color", upper="circle", mar=c(1,1,1,1),
               tl.pos="lt", diag="n", title = "Correlation Matrix Visualization",
               order="hclust", hclust.method="complete", tl.cex = .65, tl.col ="#656565")
```

```{r plotcolor, echo = FALSE}
mycolors = list("#66b3ff", "#ff80d5", "#70db70", "#d580ff", "#ffd633")
```

Fig 2. RPart Decision Tree
```{r plot1, cache = TRUE, fig.width=10, fig.height=8, fig.align = 'center'}
rpart.plot(rpart1, main="Decision Tree (rpart)", type = 1, extra=0, cex = NULL, 
           tweak = 1, fallen.leaves = FALSE, shadow.col = "#e0e0e0", box.palette = mycolors)
```

Fig 3. Random Forest OOB Error Rate
```{r rf1, fig.width=7, fig.height=4}
plot(gg_error(rfm)) + theme_bw() + scale_y_continuous(limits=c(0,.05)) + ggtitle("OOB Error Rate (Random Forest)") + geom_line(size=.75)
```


Fig 4. Variables of Importance (Random Forest)
```{r rf2, fig.width=7, fig.height=4}
vimp <- varImp(rfm); vimp <- cbind(measure = rownames(vimp), vimp)
vimp <- arrange(vimp, desc(Overall))
vimp$measure <- factor(vimp$measure, levels = vimp$measure)
ggplot(vimp[1:12,], aes(measure, Overall)) + theme_bw() +
        geom_bar(stat = "identity", fill = "#ff4d94", alpha = 0.8) +
        theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
        xlab("Measure") + 
        ggtitle("Top 12 Variables of Importance (Random Forest)")
```

Fig 5. Variables of Importance (GBM)
```{r gbm22, fig.width=7, fig.height=4}
vimp2 <- head(summary(gbm1, plotit = FALSE), 12)
vimp2$var <- factor(vimp2$var, levels = vimp2$var)
ggplot(vimp2, aes(var, rel.inf)) + theme_bw() +
        geom_bar(stat = "identity", fill = "#b366ff", alpha = 0.8) +
        theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
        xlab("Measure") + 
        ggtitle("Top 12 Variables of Importance (GBM)")
```

**Reference:**   
Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. (2013) Qualitative Activity Recognition of Weight Lifting Exercises. *Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13)* . Stuttgart, Germany: ACM SIGCHI






